\textbf{Too Beautiful to be Fake: Attractive Faces are Less Likely to be Judged as Artificially Generated}

### 2022-09-17 submission/manuscript.Rmd
address stephanie's points

  - name          : "Dominique Makowski"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "HSS 04-18, 48 Nanyang Avenue, Singapore"
    email         : "dom.makowski@gmail.com"
    orcid         : 0000-0001-5375-9967
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Data curation"
      - "Formal Analysis"
      - "Funding acquisition"
      - "Investigation"
      - "Methodology"
      - "Project administration"
      - "Resources"
      - "Software"
      - "Supervision"
      - "Validation"
      - "Visualization"
      - "Writing – original draft"
  - name          : "An Shu Te"
    affiliation   : "1"
    orcid         : 0000-0001-8008-2824
    role:
      - "Project administration"
      - "Resources"
      - "Investigation"
      - "Writing – original draft"
  - name          : "Stephanie Kirk"
    affiliation   : "1"
    orcid         : 0000-0002-9312-5552
    role:
      - "Project administration"
      - "Resources"
      - "Writing – original draft"
  - name          : "Ngoi Zi Liang"
    affiliation   : "1"
    role:
      - "Project administration"
      - "Resources"
      - "Writing – review & editing"
  - name          : "S.H. Annabel Chen"
    affiliation   : "1, 2, 3, 4"
    orcid         : 0000-0002-1540-5516
    role:
      - "Project administration"
      - "Supervision"
      - "Writing – review & editing"
affiliation:

@@

<!-- stakes and Post-truth era -->
For the first time in the history of humanity, technology has enabled the creation of near-perfect simulations indistinguishable from reality. These artificial yet realistic constructs permeate all areas of life through immersive works of fiction, deep fakes (real-like images and videos generated by deep learning algorithms), virtual and augmented reality (VR and AR), artificial beings (artificial intelligence "bots" with or without a physical form), fake news and skewed narratives, of which ground truth is often hard to access [@nightingale2022]. Such progress not only carries important consequences for the technological and entertainment sectors, but also for security and politics, for instance if used for propaganda and disinformation, recruitment into malevolent organizations, or religious indoctrination [@pantserev2020]. This issue is central to what has been coined the "post-truth era" [@lewandowsky2017beyond], in which the distinction (and lack thereof) between authentic and simulated objects will play a critical role.

<!-- Ambiguous stimuli and irrelevance of objective features -->
While not all simulations have achieved perfect realism [e.g., Computer Generated Images - CGI in movies often lack certain key details that makes them visually distinct from real images, @mcdonnell2010face], it is fair to assume that these technical limitations will become negligible in the near future, in particular in the field of faces generation and replacement [@nightingale2022; @tucciarelli2020; @moshel2022]. This fact, however, leads to a new issue: if real and fake stimuli cannot be distinguished based on their objective characteristics, how can we make judgments regarding their nature?

@@

<!-- Internal features -->
Moreover, evidence from research indicates that inter-individual characteristics also play a crucial role in the formation of beliefs of reality, with factors such as cognitive style, prior beliefs, and personality traits significantly impacting simulation monitoring [@bryanov2021determinants; @ecker2022psychological; @sindermann2020short]. For instance, individuals with higher levels of analytical reasoning have been found to better discriminate real from fake stimuli [@pehlivanoglu2021role; @pennycook2019lazy]. Prior knowledge or beliefs about the stimulus influences one's perception of it by biasing the attention deployment towards information in line with our expectations [@britt2019reasoned], and dispositional traits, such as high levels of narcissism and low levels of openness and conscientiousness, have been associated with greater susceptibility to fake news [@piksa2022cognitive; @sindermann2020short].

<!-- Reactive features -->
Beyond stimulus- and individual-related characteristics, evidence suggests that the interaction between the two, i.e., the subjective reaction associated with the experience of a given stimulus, contributes to simulation monitoring decisions. For instance, the intensity of experienced emotions have been shown to increase one's sense of presence in a simulated reality, i.e., the extent to which one feels like "being there", as if the movie or the VR environment was real [@sanchez2005presence; @makowski2017being], and beliefs that the stimuli were fake were found to result in emotion down-regulation [@makowski2019phenomenal; @sperduti2017distinctive]. In line with these findings, other studies on susceptibility to fake news have also found heightened stimulus emotionality to be associated  with greater belief [@martel2020reliance; @bago2022emotion]. Additionally, other factors, such as the stimulus' perceived self-relevance [@sperduti2016paradox; @goldstein2009pleasure], as well as familiarity [@begg1992dissociation], could also play a role in our processing and reaction to real as opposed to non-real material.

<!-- Attractiveness -->
AI-generated images of faces, due to their popularity as a target of CGI technology and the possibility of experimentally manipulating facial features, are increasingly used to study face processing as related to saliency or emotions, as well as to other important components of faces evaluation, such as trustworthiness or attractiveness [@sobieraj2014beautiful; @balas2017; @tsikandilakis2019beauty; @calbi2017]. Interestingly, some studies report that when the nature of the faces was ambiguous, artificially created faces that were previously rated as more attractive were judged by subjects to be less real [@tucciarelli2020]. However, as the attractiveness ratings were given by independent raters instead of the participants, the direction of the relationship between perceived realness and attractiveness cannot be concluded. To this end, @liefooghe2022faces reports that attractiveness ratings differed significantly between participants who were told that the faces were AI-generated from those who had no prior knowledge. Whereas this line of evidence suggests that beliefs of reality have an effect on attractiveness ratings of faces, the opposite question, whether attractiveness could drive beliefs of reality, has received little attention to date.

<!-- Goals and hypotheses -->
This study aims primarily at exploring the effect of face attractiveness on simulation monitoring, i.e., on the beliefs that an image is real or artificially generated. Based on the embodied reality theory [outlined in @makowski2018cognitive; @makowski2019phenomenal], which suggests that salient and emotional stimuli are perceived to be more real, we hypothesize a quadratic relationship between the perceived realness and attractiveness: faces rated as highly attractive or unattractive will more likely believed to be real. Moreover, we expect a similar relationship with trustworthiness ratings given its well-established link with attractiveness [@liefooghe2022faces; @bartosik2021you; @garrido2017kdef; @little2011facial], and a positive relationship with familiarity (as more familiar faces would appear as more salient, self-relevant and anchored in reality). Additionally, we will further explore the link of dispositional traits, such as personality and attitude towards AI, with inter-individual simulation monitoring tendencies. Note that the discriminative accuracy between "true" photos and "true" artificially-generated images is not relevant for this study, which focuses on the beliefs that a stimulus is real or fake, independently of its true nature.

@@

In the second part of this study, 109 photos of neutral-expression faces of real individuals from the American Multiracial Face Database (AMFD, [@chen2021broadening]) were presented to the participants for 500ms each, in a randomized order. Following each stimulus, ratings of **Attractiveness** ("I find this person attractive"), **Beauty** ("This face is good-looking"), **Trustworthiness** ("I find this person trustworthy") and **Familiarity** ("This person reminds me of someone I know") were collected using visual analog scales.

@@

We excluded 3 participants that failed 2 (out of 3) or more attention check questions. The final sample included 100 participants (Mean age = 27.9, SD = 8.5, range: [19, 66]; Sex: 48% females, 52% males).

@@

Finally, we investigated the inter-individual correlates of simulation monitoring by computing, for each participant, the proportion of faces judged as real (i.e., the overall bias towards one or the other belief), as well as the average confidence for faces judged as real, and fake. We assessed the link between these scores and dispositional traits using Bayesian correlation analysis [@makowski2019indices@; @correlationArticle].

The analysis was carried out using *R 4.2* [@RCoreTeam2022], the *tidyverse* [@wickham2019], and the *easystats* collection of packages [@insightArticle; @bayestestRArticle; @performanceArticle; @parametersArticle]. As all the details, scripts and complimentary analyses are available in open-access, the manuscript will focus on significant results.

@@

We did not find any significant relationships for familiarity adjusted for beauty and trustworthiness ($R^2_{marginal}$ = 3.0%). However, a significant positive and linear relationship was found with the confidence in faces judged as real ($\beta_{poly1} = 9.31$, $95\% CI = [3.45, 15.17]$, $z = -3.11$, $p = .002$), and a quadratic relationship for faces judged as fake ($\beta_{poly1} = -12.67$, $95\% CI = [-19.87, -5.47]$, $z = -3.45$, $p < .001$; $\beta_{poly2} = 8.14$, $95\% CI = [0.01, 16.28]$, $z = 1.96$, $p = .05$), for males only, suggesting that faces judged as real with more confidence when they are familiar, and judged as fake with less confidence when they are of not familiar or highly familiar.

@@

Notably, despite all the facial images being of real individuals retrieved from the same database, our results found every participant easily believed (to high degrees of confidence) that a significant proportion of them were fake. While this result is unsurprising, it is nevertheless a testimony to the existing expectations regarding CGI technology, as well as to the volatility of our sense of reality. In fact, stimuli-related and participant-related characteristics accounted for less than 20% of the variance in beliefs, suggesting that a large part of it is associated with other subjective processes.

@@

Although attractiveness did not seem to be the primary drive underlying simulation monitoring of facial images in the present study, it does nonetheless display an association. Furthermore, the positive quadratic relationship found between beliefs of realness and attractiveness for female participants is aligned with our hypothesis, that faces which are highly salient (i.e., perceived to be very attractive or very unattractive) are judged to be more real. Alternatively, the positive linear relationship between beauty and attractiveness, and beliefs of realness observed for males could have an evolutionary basis. In particular, males are widely postulated to place more importance toward facial attractiveness as a sign of reproductive potential, relative to females, who tend to value characteristics signaling resource acquisition capabilities [@qi2022gender]. As such, males could have perceived highly attractive and beautiful faces to be more real since such attractiveness act as salient cues of biological quality [@fink2006facial]. Therefore, our study suggests a more complex relationship between stimulus attractiveness, demographic variables and perceptions of reality that has not been examined in the literature.

>>>>

  - name          : "Dominique Makowski"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "HSS 04-18, 48 Nanyang Avenue, Singapore"
    email         : "dom.makowski@gmail.com"
    orcid         : 0000-0001-5375-9967
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Data curation"
      - "Formal Analysis"
      - "Funding acquisition"
      - "Investigation"
      - "Methodology"
      - "Project administration"
      - "Resources"
      - "Software"
      - "Supervision"
      - "Validation"
      - "Visualization"
      - "Writing – original draft"
  - name          : "An Shu Te"
    affiliation   : "1"
    orcid         : 0000-0001-8008-2824
    role:
      - "Project administration"
      - "Resources"
      - "Investigation"
      - "Writing – original draft"
  - name          : "Stephanie Kirk"
    affiliation   : "1"
    orcid         : 0000-0002-9312-5552
    role:
      - "Project administration"
      - "Resources"
      - "Writing – original draft"
  - name          : "Ngoi Zi Liang"
    affiliation   : "1"
    role:
      - "Project administration"
      - "Resources"
      - "Writing – review & editing"
  - name          : "Panagiotis Mavros"
    affiliation   : "2"
    orcid         : 0000-0002-1540-5516
    role:
      - "Supervision"
      - "Writing – review & editing"
  - name          : "S.H. Annabel Chen"
    affiliation   : "1, 3, 4, 5"
    orcid         : 0000-0002-1540-5516
    role:
      - "Project administration"
      - "Supervision"
      - "Writing – review & editing"
affiliation:

@@

<!-- stakes and Post-truth era -->
For the first time in human history, technology has enabled the creation of near-perfect simulations indistinguishable from reality. These artificial, yet realistic constructs permeate all areas of life through immersive works of fiction, deep fakes (real-like images and videos generated by deep learning algorithms), virtual and augmented reality (VR and AR), artificial beings (artificial intelligence "bots" with or without a physical form), fake news and skewed narratives, of which ground truth is often hard to access [@nightingale2022]. Such progress not only carries important consequences for the technological and entertainment sectors, but also for security and politics - for instance if used for propaganda and disinformation, recruitment into malevolent organizations, or religious indoctrination [@pantserev2020]. This issue is central to what has been coined the "post-truth era" [@lewandowsky2017beyond], in which the distinction (and lack thereof) between authentic and simulated objects will play a critical role.

<!-- Ambiguous stimuli and irrelevance of objective features -->
While not all simulations have achieved perfect realism [e.g., Computer Generated Images - CGI in movies often lack certain key details that makes them visually distinct from real images, @mcdonnell2010face], it is fair to assume that these technical limitations will become negligible in the near future, in particular in the field of face generation and replacement [@nightingale2022; @tucciarelli2020; @moshel2022]. This fact, however, leads to a new issue: if real and fake stimuli cannot be distinguished based on their objective characteristics, how can we make judgments regarding their nature?

@@

<!-- Internal features -->
Moreover, evidence from research indicates that inter-individual characteristics also play a crucial role in the formation of beliefs of reality, with factors such as cognitive style, prior beliefs, and personality traits significantly impacting simulation monitoring [@bryanov2021determinants; @ecker2022psychological; @sindermann2020short]. For instance, individuals with higher levels of analytical reasoning have been found to better discriminate real from fake stimuli [@pehlivanoglu2021role; @pennycook2019lazy]. Prior knowledge or beliefs about the stimulus influences one's perception of it by biasing the attention deployment towards information that is in line with our expectations [@britt2019reasoned]. Furthermore, dispositional traits, such as high levels of narcissism and low levels of openness and conscientiousness, have been associated with greater susceptibility to fake news [@piksa2022cognitive; @sindermann2020short].

<!-- Reactive features -->
Beyond stimulus- and individual-related characteristics, evidence suggests that the interaction between the two (i.e., the subjective reaction associated with the experience of a given stimulus), contributes to simulation monitoring decisions. For instance, the intensity of experienced emotions have been shown to increase one's sense of presence - the extent to which one feels like "being there", as if the object of experience was real - when engaged in a fictional movie or a VR environment [@sanchez2005presence; @makowski2017being]. Conversely, beliefs that emotional stimuli were fake (e.g., that emotional scenes were not authentic but instead involved actors and movie makeup) were found to result in emotion down-regulation [@makowski2019phenomenal; @sperduti2017distinctive]. In line with these findings, studies on susceptibility to fake news have also found heightened stimulus emotionality to be associated  with greater belief [@martel2020reliance; @bago2022emotion]. Additionally, other factors, such as the stimulus' perceived self-relevance [@sperduti2016paradox; @goldstein2009pleasure], as well as familiarity [@begg1992dissociation], could also play a role in our processing and reaction to real as opposed to non-real material.

<!-- Attractiveness -->
AI-generated images of faces, due to their popularity as a target of CGI technology and the possibility of experimentally manipulating facial features, are increasingly used to study face processing as related to saliency or emotions, as well as to other important components of faces evaluation, such as trustworthiness or attractiveness [@sobieraj2014beautiful; @balas2017; @tsikandilakis2019beauty; @calbi2017]. Interestingly, some studies report that when the nature of the faces was ambiguous, artificially created faces that were previously rated as more attractive were judged by subjects to be less real [@tucciarelli2020]. However, as the attractiveness ratings were given by independent raters instead of the participants, the direction of the relationship between perceived realness and attractiveness cannot be concluded. To this end, @liefooghe2022faces reports that attractiveness ratings differed significantly between participants who were told that the faces were AI-generated from those who had no prior knowledge. Whereas this line of evidence suggests that reality beliefs have an effect on face attractiveness ratings, the opposite question, whether attractiveness could drive simulation monitoring, has received little attention to date.

<!-- Goals and hypotheses -->
This study primarily aims at exploring the effect of face attractiveness on simulation monitoring, i.e., on the beliefs that an image is real or artificially generated. Based on the embodied reality theory [outlined in @makowski2018cognitive; @makowski2019phenomenal], which suggests that salient and emotional stimuli are perceived to be more real, we hypothesize a quadratic relationship between perceived realness and attractiveness: faces rated as highly attractive or unattractive will more likely believed to be real. We expect a similar relationship with trustworthiness ratings given its well-established link with attractiveness [@liefooghe2022faces; @bartosik2021you; @garrido2017kdef; @little2011facial], and a positive relationship with familiarity (as more familiar faces would appear as more salient, self-relevant and anchored in reality). Additionally, we will further explore the link of dispositional traits, such as personality and attitude towards AI, with inter-individual simulation monitoring tendencies. Note that the discriminative accuracy between "true" photos and "true" artificially-generated images is not relevant for this study, which focuses on the beliefs that a stimulus is real or fake, independently of its true nature.

@@

In the second part of this study, 109 photos of neutral-expression faces of real individuals from the American Multiracial Face Database (AMFD, [@chen2021broadening]) were presented to the participants for 500ms each, in a randomized order. Following each stimulus, ratings of **Attractiveness** ("I find this person attractive"), **Beauty** ("This face is good-looking"), **Trustworthiness** ("I find this person trustworthy") and **Familiarity** ("This person reminds me of someone I know") were collected using visual analog scales. (**TODO: add details and justifications**)

@@

We excluded 3 participants that failed 2 (>= 66%) or more attention check questions. The final sample included 100 participants (Mean age = 27.9, SD = 8.5, range: [19, 66]; Sex: 48% females, 52% males).

@@

Finally, we investigated the inter-individual correlates of simulation monitoring by computing, for each participant, the proportion of faces judged as real (i.e., the overall bias towards one or the other belief), as well as the average confidence for faces judged as real, and fake. We assessed the link between these scores and dispositional traits using Bayesian correlation analysis [@makowski2019indices; @correlationArticle].

The analysis was carried out using *R 4.2* [@RCoreTeam2022], the *tidyverse* [@wickham2019], and the *easystats* collection of packages [@insightArticle; @bayestestRArticle; @performanceArticle; @parametersArticle]. As all the details, scripts and complimentary analyses are open-access, the manuscript will focus on significant results.

@@

We did not find any significant relationships for familiarity adjusted for beauty and trustworthiness ($R^2_{marginal}$ = 3.0%). However, a significant positive and linear relationship was found with the confidence in faces judged as real ($\beta_{poly1} = 9.31$, $95\% CI = [3.45, 15.17]$, $z = -3.11$, $p = .002$), and a quadratic relationship for faces judged as fake ($\beta_{poly1} = -12.67$, $95\% CI = [-19.87, -5.47]$, $z = -3.45$, $p < .001$; $\beta_{poly2} = 8.14$, $95\% CI = [0.01, 16.28]$, $z = 1.96$, $p = .05$), for males only, suggesting that faces are judged as real with more confidence when they are familiar, and judged as fake with less confidence when they are of not familiar or highly familiar.

@@

This study aimed at investigating the effect of facial ratings (attractiveness, beauty, trustworthiness and familiarity) on simulation monitoring, i.e., on the belief that the stimulus is artificially generated. The most striking result is that despite all the images being real faces retrieved from the same validated database, all participants, when given the information, believed (to high degrees of confidence) that a significant proportion of them were fake. This result is a testimony to both the current expectations regarding CGI technology in the population, as well as to the volatility of our sense of reality. It underlines the strong impact of prior expectations and information on reality beliefs. In fact, stimuli-related and participant-related characteristics accounted for less than 20% of the beliefs variance, suggesting that a large part of it is associated with other subjective processes.

@@

Although attractiveness did not seem to be the primary drive underlying simulation monitoring of face images, we do nonetheless report significant associations, with a different pattern depending on the participant's gender. The quadratic relationship found for female participants is aligned with our hypothesis that faces which are highly salient (i.e., perceived to be very attractive or very unattractive) are judged to be more real. Alternatively, we could interpret the mostly positive linear relationship for males under an evolutionary lens.

In particular, males are widely postulated to place more importance toward facial attractiveness as a sign of reproductive potential, relative to females, who tend to value characteristics signaling resource acquisition capabilities [@qi2022gender]. As such, males could have perceived highly attractive and beautiful faces to be more real since such attractiveness act as salient cues of biological quality [@fink2006facial]. Therefore, our study suggests a more complex relationship between stimulus attractiveness, demographic variables and perceptions of reality that has not been examined in the literature.
